A Toolbox for Manipulating and Assessing Color Palettes

### 2019-10-04 pkg/colorspace/index.Rmd
feedback from 2nd round of reviews in JSS
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@510 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

```{r hcl, echo = FALSE, fig.show = "hide", fig.width = 4.5, fig.height = 4.5, fig.align = "center", dev = "png", dpi = 48}
library("colorspace")
hclplot(qualitative_hcl(9, c = 50, l = 70))
```

>>>>

```{r, hcl, echo = FALSE, fig.show = "hide", fig.width = 4.5, fig.height = 4.5, fig.align = "center", dev = "png", dpi = 48}
library("colorspace")
hclplot(qualitative_hcl(9, c = 50, l = 70))
```

### 2019-10-04 pkg/colorspace/paper/Toolbox/article.Rmd
feedback from 2nd round of reviews in JSS
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@510 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

Color is an integral and omnipresent element of many statistical graphics and data
visualizations. Therefore, colors should be carefully chosen to support all viewers in
accessing the information displayed [@color:Tufte:1990; @color:Brewer:1999; @color:Ware:2004; @color:Wilkinson:2005; @color:Wilke:2019].
However, until relatively recently many software packages have been using color
palettes derived from simple RGB (red-green-blue)
color combinations such as the RGB "rainbow" (or "jet") color palette
with poor perceptual properties. See @color:Hawkins+McNeall+Stephenson:2014
and @color:Stauffer+Mayr+Dabernig:2015 and the references therein for an overview.

@@

The remainder of the paper is organized as follows:
Section\ \ref{sec:tour} gives a first overview of the package's "look & feel" and the general workflow.
Section\ \ref{sec:color_spaces} summarizes the \proglang{S}4 color space classes and methods in the package.
Section\ \ref{sec:hcl_palettes} introduces the extensible collection of HCL-based palettes along with their construction details.
Section\ \ref{sec:palette_visualization} presents the toolbox for palette visualization and assessment.
Section\ \ref{sec:color_vision_deficiency} discusses the implemented techniques for color vision deficiency emulation
that help assess the suitability of colors for colorblind viewers.
Section\ \ref{sec:hclwizard} briefly highlights the interactive color apps from the package.
Finally, some further color manipulation utilities are highlighted in Section\ \ref{sec:manipulation_utilities}
before Section\ \ref{sec:summary} concludes the paper.

@@

To inspect the HCL parameter combinations for a specific palette simply include the `palette` name
where upper- vs.\ lower-case, spaces, etc.\ are ignored for matching the label, i.e.,
`"set2"` matches `"Set 2"`:

@@

As the properties of a palette in terms of the perceptual dimensions _hue_,
_chroma_, and _luminance_ are not always clear from looking just at color
swatches or (statistical) graphics based on these palettes, the `specplot()`
function provides an explicit display for the coordinates of the HCL trajectory associated
with a palette. This can bring out clearly various aspects, e.g.,
whether hue is constant, chroma is monotonic or triangular, and whether
luminance is approximately constant (as in many qualitative palettes),
monotonic (as in sequential palettes), or diverging.

@@

- Working best for fewer colors (e.g., bar, pie, scatter, lines, ...) vs.\ many colors (e.g., heatmap, perspective, ...).
- Intended for categorical data (e.g., bar, pie, ...) vs.\ continuous numeric data (e.g., heatmap, perspective, ...).
- Shading areas (e.g., map, bar, pie, ...) vs.\ shading points or lines (scatter, lines).

@@

There are other \proglang{R} packages that can complement the palettes provided by \pkg{colorspace}.
\pkg{Polychrome} [@color:Polychrome; @color:Polychrome2] implements strategies
for qualitative palettes with many "categories". While the qualitative palettes
in Section\ \ref{sec:hcl_palettes} yield only about 6--8 clearly distinguishable
colors due to the fixed chroma and luminance, \pkg{Polychrome} relaxes this
restriction and can thus find a larger number of colors in CIELUV space that
are spaced as far apart as possible. It also has several functions for visualizing
color palettes in different ways. 
The palette collection packages \pkg{pals} [@color:pals] and \pkg{paletteer} [@color:paletteer]
also provide a wide range of prespecified palettes, including some qualitative
schemes with many categories. Note that the palettes are quite diverse, though,
and not all of them are equally suitable for coding qualitative information.
The visualization functions in \pkg{colorspace} from Section\ \ref{sec:palette_visualization}
may be helpful in assessing their properties.
\pkg{roloc} [@color:roloc; @color:roloc2] also provides color conversions,
not between numeric color spaces, but rather from numeric color spaces to English color names.

In addition to the \proglang{R} version of \pkg{colorspace}, a
\proglang{Python}\ 2/\proglang{Python}\ 3 re-implementation is in beta
and available at <https://github.com/retostauffer/python-colorspace>.
In the paper we focus on the more mature \proglang{R} implementation but
replication materials for most examples are not only provided for both \proglang{R}
and \proglang{Python}.

@@

>>>>

Color is an integral element of many statistical graphics and data
visualizations. Therefore, colors should be carefully chosen to support all viewers in
accessing the information displayed [@color:Tufte:1990; @color:Brewer:1999; @color:Ware:2004; @color:Wilkinson:2005; @color:Wilke:2019].
However, until relatively recently many software packages have been using color
palettes derived from simple RGB (red-green-blue)
color combinations such as the RGB "rainbow" (or "jet") color palette
with poor perceptual properties. See @color:Hawkins+McNeall+Stephenson:2014
and @color:Stauffer+Mayr+Dabernig:2015 and the references therein for an overview.

@@

The remainder of the paper is organized as follows:
Section\ \ref{sec:tour} gives a first overview of the package's "look & feel" and the general workflow.
Section\ \ref{sec:color_spaces} summarizes the \proglang{S}4 color space classes and methods in the package.
Section\ \ref{sec:hcl_palettes} introduces the extensible collection of HCL-based palettes along with their construction details.
Section\ \ref{sec:palette_visualization} presents the toolbox for palette visualization and assessment.
Section\ \ref{sec:color_vision_deficiency} discusses the implemented techniques for color vision deficiency emulation
that help assess the suitability of colors for colorblind viewers.
Section\ \ref{sec:hclwizard} briefly highlights the interactive color apps from the package.
Some further color manipulation utilities are highlighted in Section\ \ref{sec:manipulation_utilities}
before Section\ \ref{sec:summary} concludes the paper.

@@

To inspect the HCL parameter combinations for a specific palette simply include the `palette` name
where upper- vs.\ lower-case, spaces, etc.\ are ignored for matching the label, e.g.,
`"set2"` matches `"Set 2"`:

@@

As the properties of a palette in terms of the perceptual dimensions _hue_,
_chroma_, and _luminance_ are not always clear from looking just at color
swatches or (statistical) graphics based on these palettes, the `specplot()`
function provides an explicit display for the coordinates of the HCL trajectory associated
with a palette. This can bring out clearly various aspects, e.g.,
whether hue is constant, whether chroma is monotonic or triangular, and whether
luminance is approximately constant (as in many qualitative palettes),
monotonic (as in sequential palettes), or diverging.

@@

- Working best for fewer colors (e.g., bar, pie, scatter, lines, ...) vs.\ many colors (e.g., heatmap, perspective, ...).
- Intended for categorical data (e.g., bar, pie, ...) vs.\ continuous numeric data (e.g., heatmap, perspective, ...).
- Shading areas (e.g., map, bar, pie, ...) vs.\ coloring points or lines (scatter, lines).

@@

There are other \proglang{R} packages that can complement the palettes provided by \pkg{colorspace}.
\pkg{Polychrome} [@color:Polychrome; @color:Polychrome2] implements strategies
for qualitative palettes with many "categories". While the qualitative palettes
in Section\ \ref{sec:hcl_palettes} yield only about 6--8 clearly distinguishable
colors due to the fixed chroma and luminance, \pkg{Polychrome} relaxes this
restriction and can thus find a larger number of colors in CIELUV space that
are spaced as far apart as possible. It also has several functions for visualizing
color palettes in different ways. 
The palette collection packages \pkg{pals} [@color:pals] and \pkg{paletteer} [@color:paletteer]
also provide a wide range of prespecified palettes, including some qualitative
schemes with many categories. Note that the palettes are quite diverse, though,
and not all of them are equally suitable for coding qualitative information.
The visualization functions in \pkg{colorspace} from Section\ \ref{sec:palette_visualization}
may be helpful in assessing their properties.
\pkg{roloc} [@color:roloc2; @color:roloc] also provides color conversions,
not between numeric color spaces, but rather from numeric color spaces to English color names.

In addition to the \proglang{R} version of \pkg{colorspace}, a
\proglang{Python}\ 2/\proglang{Python}\ 3 re-implementation is in beta
and available at <https://github.com/retostauffer/python-colorspace>.
In the paper we focus on the more mature \proglang{R} implementation but
replication materials for most examples are also available for
\proglang{Python}.

@@

# Acknowledgments {#acknowledgments .unnumbered}

The authors would like to thank the journal editors, the associate editor,
and two anonymous reviewers for their constructive and helpful feedback
that substantially improved the paper.
